Covid Fields Data Protocol

### 2023-01-04 data_cleaning/Data_cleaning_and_covariates_Jan_03.qmd
Updated geographical data

---
title: "Fields COVID-19 dataset exploration"
author: "Ariel Mundo"
format: pdf
toc: true
toc-depth: 2
editor: visual
urlcolor: blue
#bibliography: refs.bib
---

@@

```{r,outliers}
test<-clean_data %>%
  filter(age_group=="16_24" &
           #h_size==1 &
           income=="over_110000"

@@

test%>%
  ggplot(aes(x=h_size))+
  geom_histogram(stat="count")

@@

From the plot it can be seen that about 20 observations are the ones that report a household of 1. The Census data indicates that the average household size is 2.6, and that he median total income in the province is \$33,539. Therefore, this would mean that an income of \>110 could be expected from a household of 4 people and above.

@@

**Overall, we will have to weight the data because it is not representative of the population distribution in Ontario**.

@@

p1<-clean_data%>%
  select(date,city)%>%
  mutate(location=as.character(case_when(city=="Toronto"~"Toronto",
                            city!="Toronto"~ "All other cities"
                          )))%>%
  group_by(date,location)%>%
  summarize(Freq=n())%>%
  ggplot(aes(x=date, y=Freq,fill=factor(location)
                ))+
  geom_bar(position="fill",stat="identity",show.legend = TRUE)+
  theme_classic()+
  ylab("Percentage")+
  theme(legend.title=element_blank())+
  scale_x_date(date_labels = "%b-%Y")+
  thm1

# p2<-clean_data%>%
#   select(date,age_group)%>%
#   mutate(ag=as.numeric(case_when(age_group=="16_24"~1,
#                             age_group=="25_34"~ 2,
#                             age_group=="35_44"~ 3,
#                             age_group=="45_54"~ 4,
#                             age_group=="55_64"~ 5,
#                             age_group=="65_and_over"~6
#                             )))%>%
#   group_by(date,ag)%>%
#   summarise(n = n()) %>%
#   mutate(perc = n / sum(n)
#          )%>%
#   ggplot(aes(x=date,y=perc,color=factor(ag)))+
#   geom_point()+
#   geom_line(aes(x=date,y=perc,group=factor(ag)))+
#   facet_wrap(~ag)+
#   scale_x_date(date_breaks = "2 months")+
#   ggtitle("Age groups")+
#   theme_classic()+
#   thm1

p1

@@

The proportion of responses from Toronto is somewhat uniform across time, around 20%-25%. The Census data indicates that Toronto has around 38% of the total population province (5,433,590/14,223,942). This would mean that a correction would be needed also for the proportion of answers. Note that I only analyzed the proportion of answers from Toronto as it is the major urban centre of the province, but also because the dataset contains at least 500 different city entries. I did not explore geographically if some of these cities lie within what is considered Toronto (the dataset is also not explicit as the method used to assign the cities).

@@

However, when I plotted the proportion of answers at each time point among groups, the distribution looks rather different:

@@

p2<-clean_data%>%
  select(date,age_group)%>%
   mutate(age_group_ord = fct_relevel(age_group,
                                  "65_and_over",
                                  "55_64",
                                  "45_54",
                                  "35_44",
                                  "25_34",
                                  "16_24"))%>%
  group_by(date,age_group_ord)%>%
  summarize(Freq=n())%>%
  ggplot(aes(x=date, y=Freq,fill=age_group_ord))+
  geom_bar(position="fill",stat="identity",show.legend = TRUE)+
  theme_classic()+
  # scale_fill_discrete(name="Age group",
  #                        labels=c("16-24", 
  #                                 "25-34",
  #                                 "35-44",
  #                                 "45-54",
  #                                 "55-64",
  #                                 "65 and over"))+
  ylab("Percentage")+
  scale_y_continuous(breaks = seq(0, 1, by = 0.1))+
  theme(legend.title = element_blank())+
   scale_x_date(date_labels = "%b-%Y")+
  thm2

p2

# a4<-clean_data%>%
#   group_by(age_group)%>%
#   count(date,first_dose)%>%
#   ggplot(aes(x=date,y=n,color=first_dose))+
#   geom_point(show.legend = FALSE)+
#   geom_line(aes(x=date,y=n,group=first_dose),show.legend = TRUE)+
#   facet_wrap(~age_group)+
#   ggtitle("First dose")+
#   theme_classic()+
#   thm1
# 
# a5<-clean_data%>%
#   group_by(age_group)%>%
#   count(date,second_dose)%>%
#   ggplot(aes(x=date,y=n,color=second_dose))+
#   geom_point()+
#   geom_line(aes(x=date,y=n,group=second_dose))+
#   facet_wrap(~age_group)+
#   ggtitle("Second dose")+
#   theme_classic()
# 
# layout<-"AAAAA
#          AAAAA
#          AAAAA
#          AAAAA
#          BBBBB
#          BBBBB
#          BBBBB
#          BBBBB"
# 
# a4+a5+plot_layout(design = layout)
# 
# 
# 
# p3<-clean_data%>%
#   select(date,age_group)%>%
#   mutate(location=as.character(case_when(city=="Toronto"~"Toronto",
#                             city!="Toronto"~ "All other cities"
#                           )))%>%
#   group_by(date,location)%>%
#   summarize(Freq=n())%>%
#   ggplot(aes(x=date, y=Freq,fill=factor(location)
#                 ))+
#   geom_bar(position="fill",stat="identity",show.legend = TRUE)+
#   ggtitle("Geographical location of survey answers")+
#   theme_classic()+
#   ylab("Percentage")+
#   theme(legend.title=element_blank())+
#   scale_x_date(date_labels = "%b-%Y")+
#   thm1

@@

It can be seen from the graph that the trends in answers by group age do not follow the overall distribution from the province. In particular, the 16-24 group represents in most cases at least 20% of the answers, whereas the 35-44 group is less than 10% in some cases.

@@

The final aspect I wanted to explore was the income distribution on the responses over time. In this regard the question from the survey was "What is your household annual income?". To compare the results of this answer, I used the "Household total income groups in 2015 for private households" from the Census data. The information from the census appears next.

@@

p3<-clean_data%>%
  select(date,income)%>%
  mutate(income_ord = fct_relevel(income,
                                  "over_110000",
                                  "90000_109999",
                                  "60000_89999",
                                   "40000_59999",
                                  "25000_39999",
                                     "15000_24999",
                                   "under_15000"))%>%
  group_by(date,income_ord)%>%
  summarize(Freq=n())%>%
  ggplot(aes(x=date, y=Freq,fill=income_ord))+
  geom_bar(position="fill",stat="identity",show.legend = TRUE)+
  theme_classic()+
  ylab("Percentage")+
  scale_y_continuous(breaks = seq(0, 1, by = 0.1))+
  theme(legend.title = element_blank())+
   scale_x_date(date_labels = "%b-%Y")+
  thm3

p3

@@

# Conclusions

Overall, it seems that poststratification corrections would be needed for the dataset with regard to race, ethnicity, age groups

>>>>

---
title: "Fields COVID-19 dataset exploration"
author: "Ariel Mundo"
format: pdf
toc: true
toc-depth: 2
editor: source
urlcolor: blue
#bibliography: refs.bib
---

@@

```{r,outliers}
outliers<-clean_data %>%
  filter(age_group=="16_24" &
           h_size==1 &
           income=="over_110000"

@@

outliers%>%
  ggplot(aes(x=h_size))+
  geom_histogram(stat="count")

#From the plot it is about 21 entries that are outliers. Now, remove these outliers

clean_data<-anti_join(clean_data,outliers)

@@

@@

**Overall, we weighted the data because it is not representative of the population distribution in Ontario**.

@@

p1<-clean_data%>%
  select(city)%>%
  mutate(location=as.character(case_when(city=="Toronto"~"Toronto",
                            city!="Toronto"~ "All other cities"
                          )))

p1$location<-as.factor(p1$location)

ggplot(p1 %>% count(location) %>%    # then count number in each group
         mutate(pct=n/sum(n),               # Calculate percent within each group
                ypos = 200),  # Calculate label positions
       aes(location, n, fill=location)) +
  geom_bar(stat="identity") +
  geom_text(aes(label=paste0(sprintf("%1.1f", pct*100),"%"), y=ypos))+
  theme_classic()+
  thm2

@@

The proportion of responses from Toronto is 30.9% of the total. The Census data indicates that Toronto has around 38% of the total population province (5,433,590/14,223,942). This would mean that a correction would be needed also for the proportion of answers.

Additionally, the data needs to be cleaned as some of the responses do not have an assigned city, and the entry only reads "None". These entries will be removed as they lack geographical information.

```{r, remove-entries}

clean_data<-clean_data%>%
  filter(city!="None")

@@

However, when plotted the proportion of answers at each time point among groups, the distribution looks rather different.

@@

p2<-clean_data%>%
  select(age_group)%>%
   mutate(age_group_ord = fct_relevel(age_group,
                                      "16_24",
                                      "25_34",
                                      "35_44",
                                      "45_54",
                                      "55_64",
                                  "65_and_over"
                                          ))

  ggplot(p2%>%count(age_group_ord) %>%    # G count number in each group
         mutate(pct=n/sum(n),               # Calculate percent within each group
                ypos = 200),  # Calculate label positions
       aes(age_group_ord, n, fill=age_group_ord)) +
  geom_bar(stat="identity") +
  geom_text(aes(label=paste0(sprintf("%1.1f", pct*100),"%"), y=ypos))+
  theme_classic()+
  thm1

@@

It can be seen from the graph that the proportion of answers by group age do not follow the overall distribution from the province.

@@

The final aspect I wanted to explore was the income distribution. In this regard the question from the survey was "What is your household annual income?". To compare the results of this answer, I used the "Household total income groups in 2015 for private households" from the Census data. The information from the census appears next.

@@

p3<-clean_data%>%
  select(date,income)%>%
  mutate(income_ord = fct_relevel(income,
                                  "under_15000",
                                  "15000_24999",
                                  "25000_39999",
                                  "40000_59999",
                                  "60000_89999",
                                  "90000_109999",
                                  "over_110000"
                                ))

ggplot(p3%>% count(income_ord)%>%  # count number in each group
         mutate(pct=n/sum(n),               # Calculate percent within each group
                ypos = 200),  # Calculate label positions
       aes(income_ord, n, fill=income_ord)) +
  geom_bar(stat="identity") +
  geom_text(aes(label=paste0(sprintf("%1.1f", pct*100),"%"), y=ypos))+
  theme_classic()+
  thm3

@@

# Joining the municipalities data

We have obtained and cleaned the data of the municipalities from the province of Ontario website, and cleaned the dataset to obtain the city names and geographical locations. Further details can be found in `municipalities.qmd` in the `data_cleaning` directory. Next, we will join and match the geographical location from the municipalities dataset to the clean dataset we have obtained in this document so far.

```{r,join-datasets}

municipalities<-read.csv(here("data","municipalities_clean.csv"))

municipalities$Municipal.status<-as.factor(municipalities$Municipal.status)

municipalities$Geographic.area<-as.factor(municipalities$Geographic.area)

municipalities$city<-as.factor(municipalities$city)

clean_data<-left_join(clean_data,municipalities,by="city")

Need to explore which entries were left without a geographical region:

```{r, missing-geog-area}

test<-clean_data %>%
  filter(is.na(Geographic.area))%>%
  distinct(city)

There are 756 observations that did not get a geographical region, and they correspond to 177 unique cities. These will be exported to a csv file in order to manually write their geographical areas following the Association of Municipalities of Ontario divisions, and using Wikipedia to check the status of each municipality.

```{r,write-file}
#| echo: false

write.csv(test,here("data","missing_municipalities.csv"),row.names = FALSE)

The code chunk above was ran once to get the names of the municipalities. After manually entering the geographic areas, the file was saved as `missing_municipalities_updated.csv`, and this will be the file to be used for the next steps.

Note: there is one city "Kinburn", but there are two communities with such name, one in Huron County and one in Carelton County, for the time being Huron County is assigned, will check later if is better to remove it.

Also one occurence of "Sydenham", which can be a ward in Kingston or a community in Frotenac. Will assign Frotenac for the time being.

## Merging missing municipalities and geographic area names

After manually assigning the geographical regions to the municipalities that were missing (which can be found in `missing_municipalities.csv` in the `data` directory), these will be merged as the dataset `geographic_areas.csv` (also in `data`), which contains the titles for each region (e.g., "County", "Region").

```{r,merge-datasets-2}

### 2023-01-04 protocol/Background_and_methods_Dec_30.qmd
Updated geographical data

| Variable             | Values                                                                                                                          |
|:---------------------|---------------------------------------------------------------------------------------------------------------------------------|
| Age group            | 15-24,25-34,35-44,45-54,55-64, 65+                                                                                              |
| Income bracket (CAD) | \<15,000, 15,000-24,999, 25,000-39,999, 40,000-59,999, 60,000-89,999, \>90,000                                                  |
| Race/ethnicity       | Arab/Middle Eastern, Black, East Asian/Pacific Islander, Indigenous, Latin American, Mixed, South Asian, White Caucasian, other |
| Employment status    | yes, no                                                                                                                         |

@@

For each survey participant certain data was automatically captured. This included the nearest municipality, which resulted in a total of 578 different municipalities within the dataset. Because our interest lies in exploring vaccination status within a governamental administrative unit that encompasses multiple cities, we grouped the municipalities from the dataset according to the regional government aggregations provided by the Association of Municipalities of Ontario, which groups municipalities within regions, counties, districts, and single-tiers [@ontario-municipalities].

>>>>

| Variable             | Values                                                                                                                          |
|:--------------|---------------------------------------------------------|
| Age group            | 15-24,25-34,35-44,45-54,55-64, 65+                                                                                              |
| Income bracket (CAD) | \<15,000, 15,000-24,999, 25,000-39,999, 40,000-59,999, 60,000-89,999, \>90,000                                                  |
| Race/ethnicity       | Arab/Middle Eastern, Black, East Asian/Pacific Islander, Indigenous, Latin American, Mixed, South Asian, White Caucasian, other |
| Employment status    | yes, no                                                                                                                         |

@@

For each survey participant certain data was automatically captured. This included the nearest municipality, which resulted in a total of 578 different municipalities within the dataset. Because our interest lies in exploring vaccination status within a governamental administrative unit that encompasses multiple cities, we grouped the municipalities from the dataset according to the regional government aggregations provided by the Association of Municipalities of Ontario, which groups municipalities within regions, counties, districts, and single-tiers [@ontario-municipalities]. The geographical regions were obtained from the websie of the province of Ontario [https://www.ontario.ca/page/list-ontario-municipalities#section-3](https://www.ontario.ca/page/list-ontario-municipalities#section-3)

### 2023-01-04 protocol/Background_and_methods_Dec_30.tex
Updated geographical data

\begin{document}
\maketitle
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, enhanced, interior hidden, sharp corners, frame hidden, boxrule=0pt, breakable]}{\end{tcolorbox}}\fi

@@

\hypertarget{tbl-covariates}{}
\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.1457}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.8543}}@{}}
\caption{\label{tbl-covariates}Selected socio-economic factors from the
survey}\tabularnewline
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
Variable
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Values
\end{minipage} \\
\midrule()
\endfirsthead
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
Variable
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Values
\end{minipage} \\
\midrule()
\endhead
Age group & 15-24,25-34,35-44,45-54,55-64, 65+ \\
Income bracket (CAD) & \textless15,000, 15,000-24,999, 25,000-39,999,
40,000-59,999, 60,000-89,999, \textgreater90,000 \\
Race/ethnicity & Arab/Middle Eastern, Black, East Asian/Pacific
Islander, Indigenous, Latin American, Mixed, South Asian, White
Caucasian, other \\
Employment status & yes, no \\
\bottomrule()
\end{longtable}

@@

For each survey participant certain data was automatically captured.
This included the nearest municipality, which resulted in a total of 578
different municipalities within the dataset. Because our interest lies
in exploring vaccination status within a governamental administrative
unit that encompasses multiple cities, we grouped the municipalities
from the dataset according to the regional government aggregations
provided by the Association of Municipalities of Ontario, which groups
municipalities within regions, counties, districts, and
single-tiers\textsuperscript{\protect\hyperlink{ref-ontario-municipalities}{22}}.

>>>>

\begin{document}
\maketitle
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, breakable, sharp corners, frame hidden, enhanced, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi

@@

\hypertarget{tbl-covariates}{}
\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.2083}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.7917}}@{}}
\caption{\label{tbl-covariates}Selected socio-economic factors from the
survey}\tabularnewline
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
Variable
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Values
\end{minipage} \\
\midrule()
\endfirsthead
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
Variable
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Values
\end{minipage} \\
\midrule()
\endhead
Age group & 15-24,25-34,35-44,45-54,55-64, 65+ \\
Income bracket (CAD) & \textless15,000, 15,000-24,999, 25,000-39,999,
40,000-59,999, 60,000-89,999, \textgreater90,000 \\
Race/ethnicity & Arab/Middle Eastern, Black, East Asian/Pacific
Islander, Indigenous, Latin American, Mixed, South Asian, White
Caucasian, other \\
Employment status & yes, no \\
\bottomrule()
\end{longtable}

@@

For each survey participant certain data was automatically captured.
This included the nearest municipality, which resulted in a total of 578
different municipalities within the dataset. Because our interest lies
in exploring vaccination status within a governamental administrative
unit that encompasses multiple cities, we grouped the municipalities
from the dataset according to the regional government aggregations
provided by the Association of Municipalities of Ontario, which groups
municipalities within regions, counties, districts, and
single-tiers\textsuperscript{\protect\hyperlink{ref-ontario-municipalities}{22}}.
The geographical regions were obtained from the websie of the province
of Ontario
\url{https://www.ontario.ca/page/list-ontario-municipalities\#section-3}
